Does motif participation provide unique information about species risk of extinction?

### 2021-10-01 manuscript/reorder_main_text.tex

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes ranging from 50 to 100 species (in steps of 10) and connectances ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, see~\citet{Cirtwill2021_inprep}.
        We then filtered these simulated networks and removed any network containing disconnected components
        % (species or groups of species not connected to the rest of the network) 
        or any network where any species had a shortest trophic level \textgreater6 as such high trophic levels are very uncommon in empirical food webs ~\citep{Riede2011}.
		New networks were simulated to replace any removed networks, and this process was repeated until we obtained 100 suitable networks in each combination of size and connectance.
		All networks were rendered acyclic following~\citet{Allesina2009} and topologically sorted from lowest to highest trophic level (Appendix\emph{S1}).
		These processing steps allow the strictly bottom-up calculation of probability of persistence used in the Bayesian network framework.

@@

	\subsection*{Calculating persistence}

		First, we assigned all species a baseline probability of extinction ($\pi_{base}$) -- the risk of going extinct given factors not related to the food web itself. 
		For consumer species, this also represents extinction despite all resources being present. 
		Such events can, for example, reflect diseases or stochastic extinctions of small populations.

		As well as this baseline scenario, we simulated a set of scenarios where primary producers had a higher baseline probability of extinction ($\pi_{disturbed}$). 
		These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.

		As well as the intrinsic baseline probability of extinction, extinction risk for consumers is affected by access to resources. 
		When all resources are extant, the probability of extinction for a consumer $i$, $P(\lnot i)$, is equal to the baseline probability; $P(\lnot i|X_{1},...,X_{n}) = \pi_{base}$. 
		When all resources are extinct, the consumer will go extinct; $P(\lnot i|\lnot X_{1},...,\lnot X_{n})=1$.

		Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		We used a sigmoid non-linear functional response of consumers to the loss of resources (Appendix \emph{S2}), which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. 
		Probabilities of extinction for resource species are used when calculating probabilities for higher level species.

		Although it is possible to solve the Bayesian network exactly \citep{Eklof2013}, this is cumbersome for larger networks. 
		As numerical simulations are highly efficient and produce the same result \citep{Haussler2020}, we also use simulations here.
		To compute $P(\lnot i|f)$, we calculated the fraction of resources lost ($f$) for each consumer $i$.
		Persistence is determined by comparing $P(\lnot i|f)$ with a randomly drawn number from a uniform distribution between 0-1. 
		If this random number is lower than $P(\lnot i|f)$, the species is considered extinct. 
		The species' marginal probability of persistence is simply $P(i) = 1-P(\lnot i)$.
		To minimize noise due to the random draws, we repeat these calculations 100 000 times per web, with unique random draws when calculating $P(i)$.
		The final persistence for each species is the mean over these runs.

>>>>

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function "nichemodel" within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes ranging from 50 to 100 species (in steps of 10) and connectances ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details, see~\citet{Cirtwill2021_inprep}.
        We then filtered these simulated networks and removed any network containing disconnected components
        % (species or groups of species not connected to the rest of the network) 
        or any network where any species had a shortest trophic level \textgreater6 as such high trophic levels are very uncommon in empirical food webs ~\citep{Riede2011}.
		New networks were simulated to replace any removed networks, and this process was repeated until we obtained 100 suitable networks in each combination of size and connectance.
		All networks were rendered acyclic following~\citet{Allesina2009} and topologically sorted from lowest to highest trophic level (Appendix \emph{S1}).
		These processing steps allow the strictly bottom-up calculation of probability of persistence used in the Bayesian network framework.

@@

	\subsection*{Calculating persistence}	
    % basic idea behind BN modeling
        Bayesian Network modelling represents species and feeding interactions as the nodes and links of a network (see Appendix \emph{S2} for full details).
        Each species (node) is a Bernoulli random variable, being either extant or extinct. 
        The probability that a particular species is extinct is a function of the states of its resources.

		% probability of extinction equation
		Consider a species with $n$ resource species. 
		Before taking food-web structure into account, this species already has a baseline extinction  probability of extinction, $\pi_{base}$. 
		$\pi_{base}$ represents the risk of going extinct due to diseases, stochastic extinction of small populations, or other factors not related to the food web.
		$\pi_{base}$ may be constant, or it can vary to reflect greater sensitivity in some groups of species.
		In our simulations, we vary $\pi_{base}$ for basal resources between 0.1 and 0.5 but keep constant $\pi_{base}=0.1$ for consumers.

        If the focal species is a consumer with $n$ resources, its probability of extinction depends on $\pi_{base}$ and on the fraction $f$ of its resources that have gone extinct.
        Note that we assume that only the fraction of resources lost is important and not the identities of the missing resources.
        We then model the response of the consumer to changes in $f$ using a sigmoid non-linear function ($\alpha > 1, \beta > 1$; Appendix \emph{S2}.
		This type of response most accurately captures the secondary extinctions produced by dynamical models~\citep{Eklof2013}.
		The probability of extinction for each species $i$ can therefore be represented using the cumulative density function of a beta distribution:
		\begin{equation}
		\label{betafunc}
		P(\lnot i|f) = \pi_{base} + (1 - \pi_{base}) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}.
		\end{equation}

		If all resource species of a consumer \textit{C} are present ($f = 0$), the extinction probability will be $P(\lnot C|f) = \pi_{base}$. 
		Similarly, if all resources are extinct ($f = 1$), the extinction probability $P(\lnot C|f)$ will be equal to 1 and the consumer will go extinct.
		If the fraction $f$ is not 1 nor 0, the non-linear sigmoid curve will determine the response of the consumer, which will be more severe if the fraction resources lost is high and less severe if the fraction is small. This will in turn affect the extinction probability $P(\lnot C|f)$.
		For basal species, we assume that abiotic resources are always available and $f=0$, such that extinction depends only on $\pi_{base}$.

		% how we calculate persistences 
        Beginning with primary producers and working systematically upward through the food web, we use Equation \ref{betafunc} to calculate each species' probability of extinction $P(\lnot i|f)$.
        Using this probability, we then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Species are considered extinct if a random number drawn from a uniform distribution $[0-1]$ is less than $P(\lnot i|f)$.

        These simulated extinctions are used when calculating the fraction of resources lost $f$ for each consumer.
        This allows us to update Equation~\ref{betafunc} with information on extinctions of a focal consumer's resources and simulate bottom-up effects. As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then define probability of persistence as the fraction of simulations in which the species persists.
	    This process is highly efficient and produces the same result as exact methods of solving Bayesian networks \citep{Haussler2020}.

    % various disturbance scenarios
        In our baseline scenario, primary producers had a baseline extinction probability of $\pi_{base} = 0.1$. 
		We also simulated a set of scenarios where primary producers had a higher baseline probability of extinction ($\pi_{disturbed}$). 
		These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.18-0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.
